---
title: "Can we have a NB figure?"
output: html_notebook
---

```{r}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(ggplot2)
  library(patchwork)
  library(ComplexHeatmap)
})
theme_set(theme_bw())
```


```{r}
sce <- readRDS("~/Desktop/SCPCP000004_merged.rds")
sce_metadata <- metadata(sce)
```


```{r}
immune_celltypes <- c(
  "RBCs", 
  "CD4+ T cell" , 
  "Macrophage", 
  "cDC2/DC3", 
  "cDC", 
  "T cell", 
  "CD8+ T cell", 
  "Patrolling monocyte", 
  "Classical monocyte", 
  "Resident NK cell", 
  "Circulating NK cell", 
  "Treg", 
  "Monocyte", 
  "B cell", 
  "Plasma", 
  "cDC1", 
  "pDC", 
  ""
  )
```



```{r}
umap_df <- reducedDim(sce, "UMAP") |>
  as.data.frame() |>
  tibble::rownames_to_column("barcode")

sce_df <- colData(sce) |>
  as.data.frame() |>
  tibble::rownames_to_column("barcode") |>
  dplyr::select(barcode, library_id, infercnv_total_cnv, is_infercnv_reference, openscpca_celltype_annotation) |>
  dplyr::left_join(umap_df, by = "barcode") |>
  dplyr::mutate(
    cell_category = ifelse(openscpca_celltype_annotation == "Neuroendocrine", "malignant", "non-malignant"),
    cell_category_fine = dplyr::case_when(
      openscpca_celltype_annotation == "Neuroendocrine" ~ "malignant", 
      openscpca_celltype_annotation %in% c("Fibroblast", "Endothelial", "Schwann", "Unknown", "openscpca-excluded") ~ openscpca_celltype_annotation,
      .default = "immune"
    )
  )

rm(sce)
gc()
```

How many samples don't have inferCNV estimates?

```{r}
sce_df |>
  dplyr::filter(is_infercnv_reference) |>
  dplyr::count(library_id) |>
  dplyr::filter(n <= 100) |>
  dplyr::tally()
```

We expect that 16 samples were not processed with `inferCNV` due to an insufficient number of reference cells.
Let's confirm:

```{r}
processed_libraries <- sce_df |>
  dplyr::filter(is_infercnv_reference) |>
  dplyr::count(library_id) |>
  dplyr::filter(n > 100) |>
  dplyr::pull(library_id)

sce_df |>
  dplyr::filter(library_id %in% processed_libraries) |>
  dplyr::filter(is.na(infercnv_total_cnv))
```

Aside from these 16 libraries, there are three cells which were not annotated, so we can confirm we indeed have 26 samples with CNV annotations.

How many openscpca-excluded cells are there?

```{r}
sce_df |>
  dplyr::filter(library_id %in% processed_libraries) |>
  dplyr::filter(openscpca_celltype_annotation == "openscpca-excluded")
```
Very few!! We can definitely lump this with unknown.

```{r}
sce_df <- sce_df |>
  dplyr::mutate(cell_category = ifelse(
    cell_category == "openscpca-excluded", 
    "Unknown", 
    cell_category
  ))
```


## Let's make some UMAPs

### Colored by cell type



```{r, fig.width = 12, fig.height = 6}
category_umap <- ggplot(sce_df) + 
  aes(x = UMAP1, y = UMAP2, color = cell_category) + 
  geom_point(alpha = 0.6, size = 0.1) + 
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5))) +
  scale_color_brewer(palette = "Set1") +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(), 
    legend.position = "bottom"
  )

celltype_umap <- ggplot(sce_df) + 
  aes(x = UMAP1, y = UMAP2, color = cell_category_fine) + 
  geom_point(alpha = 0.6, size = 0.1) + 
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5))) +
  scale_color_brewer(palette = "Dark2") +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(), 
    legend.position = "bottom"
  ) + 
  labs(subtitle = "obviously needs factor/color fixing!")

category_umap + celltype_umap
```


### Colored by total CNV

We'll plot this next to a cell type one for interpretation

```{r, fig.width = 12, fig.height = 6}
infercnv_umap <- ggplot(sce_df) + 
  aes(x = UMAP1, y = UMAP2, color = infercnv_total_cnv) + 
  geom_point(alpha = 0.6, size = 0.1) + 
  scale_color_viridis_c() +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(), 
    legend.position = "bottom"
  ) 


infercnv_umap + celltype_umap
```

Looks ok.
