---
title: "Compare ESTIMATE results to consensus cell types"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
---

The purpose of this notebook is to compare ESTIMATE scores to consensus cell types.

## Setup


```{r setup}
renv::load()

library(ggplot2)
theme_set(theme_bw())
options(readr.show_col_types = FALSE)
```

### Paths

```{r paths}
data_dir <- here::here("analysis", "estimate", "data")
celltype_dir <- file.path(data_dir, "consensus-celltypes")

library_metadata_file <- here::here("s3_files", "scpca-library-metadata.tsv")
estimate_file <- here::here("analysis", "estimate", "results", "estimate_results.tsv")
immune_celltypes_groups_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-immune-cell-types.tsv"


# we'll name these by library id to start, and then swap to sample later
celltype_files <- list.files(
  celltype_dir, 
  pattern = "_processed_consensus-cell-types\\.tsv\\.gz$", 
  full.names = TRUE, 
  recursive = TRUE
) |>
    purrr::set_names(
    \(x) {stringr::str_split_i(basename(x), pattern = "_", i = 1)}
  )
```

### Read input data

```{r warning = FALSE}
estimate_df <- readr::read_tsv(estimate_file)
library_metadata_df <- readr::read_tsv(library_metadata_file)

id_df <- estimate_df |>
  dplyr::inner_join(
    library_metadata_df |>
      dplyr::filter(seq_unit %in% c("cell", "nucleus")) |>
      dplyr::select(scpca_sample_id, scpca_library_id),
    by = c("sample_id" = "scpca_sample_id")
  ) |>
  dplyr::select(project_id, sample_id, library_id = scpca_library_id) |>
  dplyr::distinct()

library_sample_c <- id_df$sample_id |>
  purrr::set_names(id_df$library_id)

celltype_data <- celltype_files |>
  # only keep libraries of interest
  purrr::keep_at(
    \(library_id) library_id %in% names(library_sample_c)
  ) |> 
  purrr::map(
    \(f) {
      readr::read_tsv(f) |>
        dplyr::select(project_id, sample_id, consensus_annotation, consensus_ontology)
  })

# update names to sample ids
sample_names <- library_sample_c[names(celltype_data)] |> unname()
names(celltype_data) <- sample_names

immune_df <- readr::read_tsv(immune_celltypes_groups_url)
```

### Define cell type groups

The first step is to group consensus cell types into "immune" and "stromal" categories to enable comparisons with ESTIMATE scores.

Let's first look at what the consensus cell types that we have for this data are.
For human readability, we'll look at the annotations, not ontologies, for this.

```{r}
all_celltypes <- celltype_data |>
  purrr::map(
    \(cell_df) dplyr::pull(cell_df, consensus_annotation) |> unique()
  ) |>
  purrr::reduce(union)

all_celltypes
```

* We can use the `immune_df` data frame to identify immune cell types to use in comparisons
* Since ESTIMATE defines stromal as non-cancer and non-immune (aka "everything else"), we'll define all non-immune and non-neural cell types as stromal cells
  * An important caveat to bear in mind that is that two of our projects, SCPCP000006 (Wilms tumor) and SCPCP000017 (osteosarcoma), will have cancer cell types with a high degree of similarity to several stromal cell types, which can make teasing these categories apart challenging.
  This is particularly relevant for mesangial cells in SCPCP000006; we are classifying these cell types as "stromal" but they may be highly similar to the cancer cells in those samples

```{r}
stromal_celltypes <- c(
  "chondrocyte", 
  "adipocyte", 
  "endothelial cell",
  "epithelial cell",              
  "smooth muscle cell", 
  "muscle cell", 
  "blood vessel endothelial cell", 
  "stromal cell", 
  "pericyte",                     
  "mesangial cell", 
  "fibroblast"
)

immune_celltypes <- immune_df$consensus_annotation
```
  
The following consensus cells are _not_ currently considered in either the immune or bulk group.

```{r}
other_celltypes <- all_celltypes[!(all_celltypes %in% c(stromal_celltypes, immune_celltypes))]
other_celltypes
```

Next, we'll calculate proportions of these cell types across samples.

```{r}
celltype_category_df <- celltype_data |>
  purrr::map(
    \(cell_df) {
        cell_df |>
        dplyr::mutate(
          estimate_category = dplyr::case_when(
            consensus_annotation %in% immune_celltypes ~ "immune",
            consensus_annotation %in% stromal_celltypes ~ "stromal", 
            .default = "other"
          )) |>
        dplyr::count(estimate_category) |>
        dplyr::mutate(frac = n/sum(n)) |>
        dplyr::select(-n)
    }
  ) |>
  purrr::list_rbind(names_to = "sample_id") |>
  # we need to add in project ids
  dplyr::inner_join(dplyr::select(id_df, project_id, sample_id)) |>
  dplyr::mutate(estimate_category = forcats::fct_relevel(estimate_category, "other", "stromal", "immune"))
```

What do we have?


```{r fig.height=12, fig.width=8}
ggplot(celltype_category_df) + 
  aes(x = sample_id, y = frac, fill = estimate_category) +
  geom_col() + 
  # bright color to help make stromal more visible
  scale_fill_manual(values = c("gray85", "navy", "dodgerblue")) + 
  facet_wrap(vars(project_id), scales = "free", ncol = 1) + 
  scale_y_continuous(expand = c(0,0)) +
  labs(y = "fraction of cells in sample") +
  theme(axis.text.x = element_blank()) 
```


* Nearly all samples across SCPCP000001, SCPCP000002, and SCPCP000006 lack "stromal cells" 
  * From the plot, it appears only 1 sample in SCPCP000001 and 1 sample in SCPCP000006 have stromal; if other samples have, it's an exceedingly small proportion
* Only SCPCP000006 and SCPCP000009 have samples with truly visible levels of "stromal cells", particularly SCPCP000009
  * It is more challenging in these cancer types (wilms and osteo, respectively) to distinguish among stromal and cancer cells get, so the stromal proportions here are not necessarily definitive

## Relationship between cell types and pseudobulk ESTIMATE scores

Next, we'll directly look at the relationships between fractions of these cell type groups and ESTIMATE scores.
We'll do this now with pseudobulk ESTIMATE scores, which is from the same underlying modality as the consensus cell types.

```{r}
wide_category_df <- celltype_category_df |>
  dplyr::mutate(estimate_category = glue::glue("celltype_{estimate_category}")) |>
  tidyr::pivot_wider(
    names_from = estimate_category, 
    values_from = frac, 
    values_fill = 0
  )
  
  
combined_wide_df <- estimate_df |>
  dplyr::filter(expression_type == "pseudobulk") |>
  dplyr::select(-expression_type) |>
  dplyr::mutate(score_type = glue::glue("estimate_score_{score_type}")) |>
   tidyr::pivot_wider(
    names_from = score_type, 
    values_from = score
  ) |>
  dplyr::left_join(wide_category_df)
```

```{r, fig.height = 3, fig.width=12}
ggplot(combined_wide_df) + 
  aes(x = estimate_score_immune, y = celltype_immune) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  facet_wrap(vars(project_id), scales = "free_y", nrow = 1) +
  labs(
    title = "Immune celltypes", 
    x = "ESTIMATE immune score", 
    y = "Immune cell fraction\nin consensus celltypes"
  )
```


```{r, fig.height = 3, fig.width=12}
ggplot(combined_wide_df) + 
  aes(x = estimate_score_stromal, y = celltype_stromal) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  facet_wrap(vars(project_id), scales = "free_y", nrow = 1) +
  labs(
    title = "Stromal celltypes", 
    x = "ESTIMATE stromal score", 
    y = "Stromal cell fraction\nin consensus celltypes"
  )
```

* Immune plots:
  * SCPCP000001, SCPCP00006, and SCPCP000017 show a positive relationship between quantities
  * There is no apparent relationship for SCPCP000002 or SCPCP000009
* Stromal plots:
  * SCPCP00006 shows a positive relationship between quantities, and SCPCP000017 _may_ be trending towards a significant positive relationship (but that's a fairly wide +/- SE)
  * There is no apparent relationship for SCPCP000001, SCPCP000002, or SCPCP000009

That said, SCPCP000006 has an outlier point that may be driving the observed positive relationships.


### Relationship between other cell types and tumor purity

In theory, the `Unknown` cells in our consensus annotation are more likely to be tumor cells.
We also excluded several cell types from analysis which might also be tumor.
Similarly in theory, the ESTIMATE score (sum of immune and stromal) is expected to be proportional to tumor purity.

Is there a relationship between these quantities?

```{r, fig.height = 3, fig.width=12, message = FALSE}
unknown_df <- celltype_data |>
  purrr::map(
    \(cell_df) {
        cell_df |>
          # group to keep these columns
          dplyr::group_by(sample_id, project_id) |>
          dplyr::summarize(
            is_unknown = sum(consensus_annotation %in% other_celltypes), 
            frac_unknown = is_unknown / dplyr::n()
          )
    }
  ) |>
  purrr::list_rbind(names_to = "sample_id") |>
  dplyr::inner_join(combined_wide_df) |>
  dplyr::mutate(estimate_score = estimate_score_stromal + estimate_score_immune)

ggplot(unknown_df) + 
  aes(x = estimate_score, y = frac_unknown) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  facet_wrap(vars(project_id), nrow = 1) +
  labs(
    x = "ESTIMATE score", 
    y = "Fraction of non-immune/non-stromal\ncell types"
  )
```

There appears to be a negative relationship for SCPCP000001 and SCPCP000017, a very weak negative relationship for SCPCP000006 (possibly outlier driven?), and no relationship for SCPCP000002 or SCPCP000009.

This may reflect that the genes in ESTIMATE do not comprehensively capture all non-cancer genes, and/or include genes that are highly expressed in some of our tumor cells.


### All projects together


We'll next visualize the same results, but pooling all projects together and with correlation coefficients (both $R$ and $\rho$).
We'll also color points based on whether the pseudobulk was derived from single-cell or single-nuclei experiment.

Again, note that score shown in the the third plot with the "other" celltypes is the ESTIMATE score, which is the sum of stromal and immune.

```{r fig.height=4, fig.width=12}
combined_wide_df <- combined_wide_df |>
  dplyr::mutate(
    seq_unit = ifelse(
      project_id %in% c("SCPCP000001", "SCPCP000002"), 
      "single-cell", 
      "single-nuclei"
    ), 
     estimate_score = estimate_score_stromal + estimate_score_immune
  )


# Helper function to make scatterplots showing correlation coefficients
make_scatterplot <- function(combined_wide_df, estimate_col, celltype_col, type, annotate_x, annotate_y) {
  r <- cor(combined_wide_df[[estimate_col]], combined_wide_df[[celltype_col]]) |> round(2)
  rho <- cor(combined_wide_df[[estimate_col]], combined_wide_df[[celltype_col]], method = "spearman") |> round(2)
  
  plot_df <- combined_wide_df |>
    dplyr::select(all_of(c(estimate_col, celltype_col, "seq_unit")))
  colnames(plot_df) <- c("estimate", "celltype", "seq_unit")
  
  ggplot(plot_df) + 
  aes(x = estimate, y = celltype) + 
  geom_point(aes(color = seq_unit)) + 
  geom_smooth(method = "lm") +
  labs(
    title = glue::glue("{type} celltypes"), 
    x = glue::glue("ESTIMATE {type} score"), 
    y = glue::glue("{type} cell fraction\nin consensus celltypes")
  ) + 
  annotate(
    geom = "text", 
    x = annotate_x, 
    y = annotate_y, 
    label = glue::glue("R = {r}\nRho = {rho}")
  )
}
patchwork::wrap_plots(
  make_scatterplot(combined_wide_df, "estimate_score_immune", "celltype_immune", "immune", -1700, 0.75),
  make_scatterplot(combined_wide_df, "estimate_score_stromal", "celltype_stromal", "stromal", -1800, 0.25),
  make_scatterplot(combined_wide_df, "estimate_score", "celltype_other", "other", -3000, 0.6), 
  guides = "collect"
)
```

We see the strongest correlation for immune celltypes as expected, this plot also emphasizes the trend that single-cell show higher fractions of immune cells than do single-nuclei samples.

## Session info

```{r}
sessionInfo()
```