---
title: "Initial exploration and comparison of pseudobulk vs. bulk expression"
author: Stephanie J. Spielman
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
---

The goal of this notebook is to compare pseudobulk and bulk calculations to determine which calculation flavors we should proceed with.
We compare two pseudobulk and two bulk measures:

* Pseudobulk:
  * `pseudobulk_deseq`: raw counts summed and normalized with `DESeq2`
  * `pseudobulk_log_counts`: raw counts summed and with a `log2(x+1)` transformation
* Bulk:
  * `bulk_counts`: Bulk raw counts normalized with `DESeq2`
  * `bulk_tpm`: TPM bulk data on a `log2` scale
  
  
We'll explore expression distributions, the relationships between bulk and pseudobulk measures, and we'll also have a brief look at  whether there are any genes with strong disagreement between bulk and single-cell. 


## Setup

```{r setup}
renv::load()

library(ggplot2)
theme_set(theme_bw())
```

### Paths

```{r paths}
data_dir <- here::here("analysis", "pseudobulk-bulk-prediction", "data")
tpm_dir <- file.path(data_dir, "tpm")
pseudobulk_dir <- file.path(data_dir, "pseudobulk")
bulk_counts_file <- file.path(data_dir, "scpca_data", "normalized_bulk_counts.rds")

tpm_files <- list.files(
  path = tpm_dir,
  full.names = TRUE,
  pattern = "-tpm\\.tsv$"
)
tpm_names <- stringr::str_split_i(basename(tpm_files), pattern = "-", i = 1)
names(tpm_files) <- tpm_names


pseudobulk_files <- list.files(
  path = pseudobulk_dir,
  full.names = TRUE,
  pattern = "-pseudobulk\\.tsv$"
)
pseudobulk_names <- stringr::str_split_i(basename(pseudobulk_files), pattern = "-", i = 1)
names(pseudobulk_files) <- pseudobulk_names

# Make sure we have the same projects, in the same order
stopifnot(
  all.equal(names(tpm_files), names(pseudobulk_files))
)
```

### Read and prepare input data

We'll make both a long and wide version of the data for convenience throughout the notebook.


```{r}
project_df_list <- purrr::map2(
  tpm_files, 
  pseudobulk_files, 
  \(tpm_file, pseudo_file) {
    
    tpm_df <- readr::read_tsv(tpm_file, show_col_types = FALSE) |>
      #  TPM needs to be in log2 space
      dplyr::mutate(expression = log2(expression))
    
    pseudo_df <- readr::read_tsv(pseudo_file, show_col_types = FALSE) 
    
    dplyr::bind_rows(
      tpm_df, 
      pseudo_df
    ) |>
      tidyr::pivot_wider(
        names_from = expression_type, 
        values_from = expression
      )
  }
)
```

We'll read in the bulk counts data and combine with the rest of the data.

```{r}
# we only want to keep these samples from the bulk counts
present_samples <- project_df_list |>
  purrr::map(
    \(df) {
      df |> 
        dplyr::pull(sample_id) |> 
        unique()
    }
  ) |>
  purrr::reduce(c)

# Make a list of data frames of bulk counts, normalized by DESeq2
bulk_df_list <- readr::read_rds(bulk_counts_file) |>
  purrr::map(
    \(df) {
      df |>
        dplyr::filter(sample_id %in% present_samples) |>
        dplyr::rename(ensembl_id = gene_id)
    }
  )

# Combine the data
project_df_list <- purrr::map2(
  bulk_df_list, 
  project_df_list, 
  \(df_counts, df_main) {
    
    df_main |>
      dplyr::left_join(df_counts, by = c("ensembl_id", "sample_id"))
  }
)
```


## Full distributions

First, we'll visualize distributions of all quantities:

```{r, fig.width = 7, fig.height = 7, warning = FALSE}
plot_df <- project_df_list |>
  purrr::list_rbind(names_to = "project_id") |> 
  tidyr::pivot_longer(
    c(-project_id, -ensembl_id, -sample_id), 
    names_to = "expression_type", 
    values_to = "expression"
  )

ggplot(plot_df) + 
  aes(x = expression, fill = expression_type) + 
  geom_density(alpha = 0.5) + 
  scale_fill_brewer(palette = "Dark2") + 
  facet_grid(
    rows = vars(expression_type), 
    cols = vars(project_id),
    scales = "free_y"
  ) +
  theme(legend.position = "none")
```

We see big spikes at zero for pseudobulk, not surprisingly.
Due to the different transformation approaches, the `pseudobulk_deseq` version has some negatives for fractional values, but the other quantities have a lower bound of zero.
All around, distributions are concetrated range from their lower bound to around 20, so it's nice to know pseudobulk and bulk are definitely on the same scale.
In the bulk counts, `SCPCP000017` seems to have a much larger peak at zero compared to other projects.

  
Let's clean up for memory here.
```{r}
rm(plot_df)
gc()
```

## Relationship between quantities

This section will look at the relationship among quantities:

* How similar are the pseudobulk measures themselves?
* How does each pseudobulk measure compare to bulk TPM?
  * For these plots, we'll bin data to see the concentration of overlapping points more easily.
* How does each pseudobulk measure compare to bulk normalized counts (not TPM)?
  * For these plots, we'll bin data to see the concentration of overlapping points more easily.
In all plots, the red line is `y=x`, and the blue line is the regression line.


### Compare pseudobulk measures

```{r fig.height=34, fig.width=8, message=FALSE, warning=FALSE}
project_df_list |>
  purrr::imap(
    \(df, project_id) {
      
      ggplot(df) + 
        aes(x = pseudobulk_deseq, y = pseudobulk_log_counts) + 
        geom_point(alpha = 0.2, size = 0.5) + 
        geom_smooth(method = "lm", linewidth = 0.5) +
        geom_abline(linewidth = 0.5, color = "red") + 
        facet_wrap(vars(sample_id), nrow = 5) +
        ggtitle(project_id) 

    }
  ) |>
  patchwork::wrap_plots(ncol = 1)
```


These quantities are exceptionally similar with these differences:

* Driven by different normalization approaches, genes with very low to zero expression
* In a handful of samples (1-2 per project), `pseudobulk_log_counts` appears to have a higher proportion of low to zero counts, and throughout has lower values than `pseudobulk_deseq`.



### Compare pseudobulk to bulk

Since these quantities here are so similar, these scatterplots will show only bulk comparisons to `pseudobulk_deseq`.

These plots will show:

* Left panel: `bulk TPM ~ pseudobulk_deseq`
* Right panel: `bulk counts ~ pseudobulk_deseq`

```{r}
# Helper function to visualize scatterplots with geom_bin_2d()
make_binned_scatterplots <- function(df, project_id, nbins, facet_rows) {
  p1 <- ggplot(df) + 
    aes(x = pseudobulk_deseq, y = bulk_tpm) + 
    geom_bin_2d(bins = nbins) + 
    geom_smooth(method = "lm", alpha = 0.8, linewidth = 0.5) +
    geom_abline(alpha = 0.8, linewidth = 0.5, color = "red") + 
    facet_wrap(vars(sample_id), nrow = facet_rows) +
    ggtitle("bulk_tpm ~ deseq") 
  
  p2 <- ggplot(df) + 
    aes(x = pseudobulk_deseq, y = bulk_counts) + 
    geom_bin_2d(bins = nbins) + 
    geom_smooth(method = "lm", alpha = 0.8, linewidth = 0.5) +
    geom_abline(alpha = 0.8, linewidth = 0.5, color = "red") + 
    facet_wrap(vars(sample_id), nrow = facet_rows) +
    ggtitle("bulk_counts ~ deseq") 


  patchwork::wrap_plots(p1, p2, ncol = 2) + patchwork::plot_annotation(title = project_id)
}
```




```{r fig.height = 12, fig.width = 12, message=FALSE, warning=FALSE}
make_binned_scatterplots(
  project_df_list$SCPCP000001, 
  project_id = "SCPCP000001",
  nbins = 15,
  facet_rows = 6
)
```



```{r fig.height = 10, fig.width = 12, message=FALSE, warning=FALSE}
make_binned_scatterplots(
  project_df_list$SCPCP000002, 
  project_id = "SCPCP000002",
  nbins = 15,
  facet_rows = 6
)
```


```{r fig.height = 12, fig.width = 12, message=FALSE, warning=FALSE}
make_binned_scatterplots(
  project_df_list$SCPCP000006, 
  project_id = "SCPCP000006",
  nbins = 15,
  facet_rows = 9
)
```


```{r fig.height = 3, fig.width = 12, message=FALSE, warning=FALSE}
make_binned_scatterplots(
  project_df_list$SCPCP000009, 
  project_id = "SCPCP000009",
  nbins = 15,
  facet_rows = 1
)
```


```{r fig.height = 8, fig.width = 12, message=FALSE, warning=FALSE}
make_binned_scatterplots(
  project_df_list$SCPCP000017, 
  project_id = "SCPCP000017",
  nbins = 40,
  facet_rows = 7
)
```

* `bulk_counts` generally has much closer to a 1:1 relationship with pseudobulk compared to `bulk_tpm`
  * This is nearly universally the case for `SCPCP000001`, `SCPCP000002`, and `SCPCP000006`, and while true for `SCPCP000009` and `SCPCP000017`, it is less pronounced
* Relationships in general are weakest for `SCPCP000017`, but the situation does look better with `bulk_counts`


#### Statistics

Let's nail this down further with correlations comparing each measure for bulk with each measure for pseudobulk.
We'll perform correlations on a per-sample basis, both parametric and non-parametric, display some correlations below both as boxplots and the full table.

```{r fig.height = 10, fig.width = 8}

# Helper function to run correlations
model_samples <- function(id, df) {
  sample_df <- df |>
    dplyr::filter(sample_id == id) 
  
  #### Bulk tpm
  df_deseq <- sample_df |>
    dplyr::filter(is.finite(pseudobulk_deseq), is.finite(bulk_tpm))
  r_deseq_tpm <- cor(df_deseq$bulk_tpm, df_deseq$pseudobulk_deseq, method = "spearman")
  rho_deseq_tpm <- cor(df_deseq$bulk_tpm, df_deseq$pseudobulk_deseq, method = "pearson")

  df_log_counts <- sample_df |>
      dplyr::filter(is.finite(pseudobulk_log_counts), is.finite(bulk_tpm))
  r_logcounts_tpm <- cor(df_deseq$bulk_tpm, df_deseq$pseudobulk_log_counts, method = "spearman")
  rho_logcounts_tpm <- cor(df_deseq$bulk_tpm, df_deseq$pseudobulk_log_counts, method = "pearson")
      
  #### Bulk counts
  df_deseq <- sample_df |>
    dplyr::filter(is.finite(pseudobulk_deseq), is.finite(bulk_counts))
  r_deseq_counts <- cor(df_deseq$bulk_counts, df_deseq$pseudobulk_deseq, method = "spearman")
  rho_deseq_counts <- cor(df_deseq$bulk_counts, df_deseq$pseudobulk_deseq, method = "pearson")

  df_log_counts <- sample_df |>
      dplyr::filter(is.finite(pseudobulk_log_counts), is.finite(bulk_counts))
  r_logcounts_counts <- cor(df_deseq$bulk_counts, df_deseq$pseudobulk_log_counts, method = "spearman")
  rho_logcounts_counts <- cor(df_deseq$bulk_counts, df_deseq$pseudobulk_log_counts, method = "pearson")
      
  # Tabulate and return some fit stats
  data.frame(
    r =      c(r_deseq_tpm,    r_logcounts_tpm,    r_deseq_counts,   r_logcounts_counts),
    rho =    c(rho_deseq_tpm,  rho_logcounts_tpm,  rho_deseq_counts, rho_logcounts_counts),
    bulk =   c("bulk_tpm",    "bulk_tpm",          "bulk_counts",    "bulk_counts"), 
    pseudo = c("deseq",       "log_counts",        "deseq",          "log_counts")
  ) |>
    tidyr::pivot_longer(
      c(r, rho), 
      names_to = "measure", 
      values_to = "correlation"
    )
}

stats_df <- project_df_list |>
  purrr::map(
    \(df) {
      
      # We need to map over sample ids now
      samples <- unique(df$sample_id)
      names(samples) <- samples
      
      fit_table <- samples |>
        purrr::map(model_samples, df) |>
        purrr::list_rbind(names_to = "sample_id")
      
      return(fit_table)

    }
  ) |>
  # now, combine all projects into a single table
  purrr::list_rbind(names_to = "project_id") |>
  dplyr::mutate(comparison = glue::glue("{bulk}-{pseudo}")) |>
  dplyr::select(-bulk, -pseudo)


ggplot(stats_df) + 
  aes(x = comparison, fill = comparison, y = correlation) + 
  geom_boxplot(linewidth = 0.25, outlier.size = 0.25)+
  facet_grid(
    rows = vars(project_id), 
    cols = vars(measure)) + 
  theme(
    legend.position = "none", 
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

* Relationships between pseudobulk and bulk counts are exceptionally similar here, and further we are generally lying along the 1:1 line.
* For all projects, the nonparametric approach yields marginally higher correlations, but often marginally
* Using a nonparametric test yields stronger correlations for `bulk_counts` not but for `bulk_tpm`
* For `SCPCP000001`, `SCPCP000002`, and `SCPCP000006`, the correlations are generally similar regardless of which quantities are being compared and which statistic is used, although there are some small differences (not likely significant)
* For `SCPCP000009` (again, only 3 samples here!) and `SCPCP000017`, `bulk_counts` tends to outperform `bulk_tpm` for either pseudobulk measure

The underlying correlations are here:


```{r}
stats_df
```



## Disagreeing expression

Based on performances of quantities above, this section considers specifically `bulk_counts` and `pseudobulk_log_counts`.

Next, we'll take a quick look at cases where one modality has zero expression and the other doesn't.
In these cases, if expression is generally high, we have evidence of disagreement/discrepancy between bulk and single-cell that may be interesting to investigate.
In this notebook, we'll just a sense of how much "there is there," and we'll leave the in-depth look into any such genes for a subsequent notebook.

In this section, we'll also use a threshold of `1e-12` for zero here.


### Bulk TPM when single-cell is zero


```{r fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
plot_df <- purrr::list_rbind(project_df_list, names_to = "project_id") |>
  dplyr::filter(
    pseudobulk_log_counts <= 1e-12, 
    bulk_counts > 1e-12
  )  
  
ggplot(plot_df) + 
  aes(x = sample_id, y = bulk_counts) + 
  geom_boxplot(outlier.size = 0.25) +
  facet_wrap(vars(project_id), ncol = 1, scale = "free") +
  theme(axis.text.x = element_blank())
```



### Single-cell when bulk TPM is zero


```{r fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
plot_df <- purrr::list_rbind(project_df_list, names_to = "project_id") |>
  dplyr::filter(
    bulk_counts <= 1e-12, 
    pseudobulk_log_counts > 1e-12
  )  
  
ggplot(plot_df) + 
  aes(x = sample_id, y = pseudobulk_log_counts) + 
  geom_boxplot(outlier.size = 0.25) +
  facet_wrap(vars(project_id), ncol = 1, scale = "free") +
  theme(axis.text.x = element_blank())
```



From both comparisons, there's a decent number of genes with high expression in one modality and essentially zero in the other. 
A more careful investigation here look into what exactly these genes are, and whether they have some biological relationship that might suggest modalities are picking up different information.



## Session info

```{r}
sessionInfo()
```