---
title: "`r glue::glue('ORA with consensus cell types: {params$project_id}')`"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  project_id: "SCPCP000001"
  summary_function: "median"
  sd_threshold: 2.5
  reps: 10000
  use_observed_only: FALSE
---

## Introduction

This notebook performs over-representation analysis (ORA) across gene sets comprised of cell-type--specific marker genes on the model residuals of `bulk ~ pseudobulk + (1+sample)` model for project `r params$project_id`.

* We use permutation testing with `r params$reps` replicates to assess significance, using an FDR correction for multiple testing. 
* We take the `r params$summary_function` of residuals across samples to have one value per gene
* Test for both bulk and pseudobulk over-representation by testing if genes outlier status is associated with being a cell type marker gene. 
  * Outliers in the bulk direction are identified as `r params$sd_threshold` standard deviations above the mean
  * Outliers in the pseudobulk are identified as `r -1*params$sd_threshold` standard deviations below the mean 

We use an odds ratio as our test statistic from this contingency table:

|                                 | Gene is outlier | Gene is not outlier |
|--------------------------------:|:---------------:|:-------------------:|
|Gene is cell type marker gene    |       A         |       B             |
|Gene is not cell type marker gene|       C         |       D             |

(Here, "is cell type marker gene" is equivalent to "in the given gene set.")

We can calculate this odds ratio as either:

* `(A/B) / (C/D)`: The odds an outlier is a cell-type marker gene compared to non-outlier
* `(A/C) / (B/D)`: The odds an cell type marker gene is an outlier compared to other genes

These are mathematically equivalent.


## Setup


```{r setup}
renv::load()

library(ggplot2)
theme_set(theme_bw())

set.seed(2025)
```

### Paths

```{r paths}
result_dir <- here::here("analysis", "pseudobulk-bulk-prediction", "results")
model_dir <- here::here(result_dir, "models")
data_dir <- here::here("analysis", "pseudobulk-bulk-prediction", "data")

model_data_file <- file.path(
  model_dir, 
  glue::glue("{params$project_id}_bulk-pseudobulk-data_threshold-0.tsv")
) 
geneset_file <- file.path(
  data_dir, 
  glue::glue("{params$project_id}_panglao-genesets.tsv")
) 

output_file <- file.path(
  result_dir, 
  glue::glue("{params$project_id}_ORA-odds-ratios.tsv")
)
if (params$use_observed_only) {
  output_file <- stringr::str_replace(
    output_file,
    ".tsv", 
    "_observed-genesets-only.tsv"
  )
}
```

### Functions

```{r}
# This function establishes a contingency table of outlier/in geneset for a given data frame of residuals
#  and calculates odds ratios in both bulk and pseudobulk directions
analyze_outliers <- function(residual_df, geneset_ensembl_ids, outlier_threshold = params$sd_threshold) {
  
  ## Positive outliers: bulk direction
  or_bulk <- residual_df |>
    dplyr::mutate(
      outlier = scaled_summary_residual > outlier_threshold,
      in_geneset = ensembl_id %in% geneset_ensembl_ids
    ) |>
    dplyr::summarize(
      odds_num = sum(in_geneset & outlier) / sum(in_geneset & !outlier),
      odds_denom = sum(!in_geneset & outlier) / sum(!in_geneset & !outlier),
      odds_ratio = odds_num / odds_denom
    ) |>
    dplyr::pull(odds_ratio)
    
    
      
  ## Negative outliers: pseudobulk direction
  or_pseudobulk <- residual_df |>
    dplyr::mutate(
      outlier = scaled_summary_residual < (-1*outlier_threshold), 
      in_geneset = ensembl_id %in% geneset_ensembl_ids
    ) |>
    dplyr::summarize(
      odds_num = sum(in_geneset & outlier) / sum(in_geneset & !outlier),
      odds_denom = sum(!in_geneset & outlier) / sum(!in_geneset & !outlier),
      odds_ratio = odds_num / odds_denom
    ) |>
    dplyr::pull(odds_ratio)
  
  data.frame(or_bulk = or_bulk, or_pseudobulk = or_pseudobulk)

}
```

## Read in data 



```{r}
model_data <- readr::read_tsv(model_data_file)
geneset_df <- readr::read_tsv(geneset_file)

if (params$use_observed_only) {
  geneset_df <- geneset_df |>
    dplyr::filter(observed_in_singlecell)
} 
geneset_list <- split(geneset_df, geneset_df$cell_type_name)
geneset_names <- names(geneset_list)
```


## Calculate odds ratios  

```{r}
# Summarize residuals to get one value per gene
summary_residual_df <- model_data |>
  dplyr::group_by(ensembl_id) |>
  # summarize based on parameter
  dplyr::summarize(
    summary_residual = do.call(params$summary_function, list(residual))
  ) |>
  # zscore for outlier detection
  dplyr::mutate(scaled_summary_residual = scale(summary_residual)) 

# Calculate ORs on real data
baseline_or_df <- geneset_list |>
  purrr::map(
    \(gs_df) analyze_outliers(summary_residual_df, gs_df$ensembl_id)
  ) |>
  purrr::list_rbind(names_to = "geneset_cell_type")
```

Here are the odds ratios calculated directly from the data:

```{r}
baseline_or_df
```

## Permutation test 

To calculate P-values here, we shuffle the residual/gene_id relationship and perform a permutation test.

```{r}
shuffled_or_df <- 1:params$reps |>
  purrr::set_names(1:params$reps) |>
  purrr::map(\(i) {
    
    shuffled_df <- data.frame(
      ensembl_id = summary_residual_df$ensembl_id, 
      scaled_summary_residual = sample(summary_residual_df$scaled_summary_residual, nrow(summary_residual_df))   
    )
    
    geneset_list |>
      purrr::map(
        \(gs_df) analyze_outliers(shuffled_df, gs_df$ensembl_id)
      ) |>
      purrr::list_rbind(names_to = "geneset_cell_type") |>
      # rename columns to with shuffled_ prefix for code clarity
      dplyr::rename(
        shuffled_or_bulk = or_bulk, 
        shuffled_or_pseudobulk = or_pseudobulk
      ) 
  }) |> purrr::list_rbind(names_to = "replicate")
```


Below, we visualize the shuffled odds ratio distributions, with a red line representing the odds ratio calculated from the real data.

```{r fig.height=10, fig.width=10}
ggplot(shuffled_or_df) + 
  aes(x = shuffled_or_bulk) + 
  geom_histogram(binwidth = 0.5, fill = "gray65") +
  facet_wrap(vars(geneset_cell_type), scales = "free_x") +
  geom_vline(
    data = baseline_or_df, 
    aes(xintercept = or_bulk), 
    color = "firebrick", 
    linewidth = 0.75
  ) + 
  labs(
    title = "Odds ratio distributions in the bulk direction", 
    x = "Shuffled bulk-direction odds ratios"
  )
```


```{r fig.height=10, fig.width=10}
ggplot(shuffled_or_df) + 
  aes(x = shuffled_or_pseudobulk) + 
  geom_histogram(binwidth = 0.5, fill = "gray65") +
  facet_wrap(vars(geneset_cell_type), scales = "free_x") +
  geom_vline(
    data = baseline_or_df, 
    aes(xintercept = or_pseudobulk), 
    color = "firebrick", 
    linewidth = 0.75
  ) + 
  labs(
    title = "Odds ratio distributions in the pseudobulk direction", 
    x = "Shuffled pseudobulk-direction odds ratios"
  )
```


Next, we can calculate P-values for each gene set.
We'll also adjust P-values for multiple testing using FDR (Benjamini-Hochberg).

```{r}
permutation_pvalues <- geneset_names |>
  purrr::set_names(geneset_names) |>
  purrr::map(
    \(geneset_name) {
      
    # This will have one row for the given geneset
    geneset_baseline_df <- baseline_or_df |>
      dplyr::filter(geneset_cell_type == geneset_name)
    
    # Define these variables for code clarity
    baseline_bulk_or <- geneset_baseline_df$or_bulk
    baseline_pseudobulk_or <- geneset_baseline_df$or_pseudobulk
      
    shuffled_or_df |>
      dplyr::filter(geneset_cell_type == geneset_name) |>
      # Calculate P-value as fraction of shuffled ORs >= baseline OR (with +1 correction on numerator & denominator to avoid p=0)
      dplyr::summarize(
        bulk = (sum(shuffled_or_bulk >= baseline_bulk_or) + 1) / (params$reps + 1), 
        pseudobulk = (sum(shuffled_or_pseudobulk >= baseline_pseudobulk_or) + 1) / (params$reps + 1)
      ) |>
      # Correct for multiple testing with FDR
      tidyr::pivot_longer(
        everything(),
        names_to = "modality", 
        values_to = "p_value"
      ) |>
      dplyr::mutate(p_adj_bh = p.adjust(p_value, method = "BH"))
  }) |>
  purrr::list_rbind(names_to = "geneset_cell_type") 

# Join the odds ratios in with their pvalues, all in long format
permutation_results <- baseline_or_df |>
  tidyr::pivot_longer(
    starts_with("or_"), 
    names_to = "modality", 
    values_to = "odds_ratio"
  ) |>
  dplyr::mutate(modality = stringr::str_remove(modality, "^or_")) |>
  # and join in the pvalues
  dplyr::left_join(permutation_pvalues)

```




How many gene sets (cell types) are significant per modality, considering the BH-adjusted p-value?

```{r}
permutation_results |>
  dplyr::filter(p_adj_bh <= 0.05) |>
  dplyr::count(modality)
```

What are those gene sets (cell types), if any?

```{r}
# wrangle a bit to make this easily human-readable
permutation_results |>
  dplyr::filter(p_adj_bh <= 0.05) |>
  dplyr::group_by(modality) |>
  dplyr::arrange(p_adj_bh, .by_group = TRUE)
```



## Export


We'll export the odds ratios and p-values to a data frame.

```{r}
readr::write_tsv(permutation_results, output_file)
```


## Session info

```{r}
sessionInfo()
```