---
title: "EPIC signature genes"
output: html_notebook
---

Previously, we observed in [`quantiseq-tumor-genes.Rmd`](quantiseq-tumor-genes.Rmd) that some of the signature genes in `quanTIseq`'s reference are recorded with different symbols in our ensembl/gene symbol mapping.
This may also be the case with `EPIC` signature genes.

If any genes present in the `EPIC` signatures, `Tref` and `Bref` are not present in our gene symbol mapping, we will need to find which ones to recode them before inference. 
The goal of this notebook is to identify any such discrepancies and determine how to recode those symbols.

## Setup

```{r setup}
renv::load()

deconv_dir <- here::here("analysis", "bulk-deconvolution")
```

First, we'll read in our ensembl/symbol mapping table and pull out our gene symbols.

```{r}
map_table <- readr::read_tsv(
  file.path(deconv_dir, "data", "reference", "ensembl_symbol.tsv")
)
gene_symbols <- map_table$gene_symbol
```


## Tref reference

Are any `Tref` genes missing from our table?

```{r}
setdiff(EPIC::TRef$sigGenes, gene_symbols)
```
This gene is an alias for `DIPK2B` (source: https://www.ncbi.nlm.nih.gov/gene/79742) which is present in our gene symbols:

```{r}
map_table |>
  dplyr::filter(gene_symbol == "DIPK2B")
```

## Bref reference


Are any `Bref` genes missing from our table?

```{r}
setdiff(EPIC::BRef$sigGenes, gene_symbols)
```
* `CECR1` is an alias for `ADA2` (source: https://www.ncbi.nlm.nih.gov/gene/51816)
* `HIST1H2BC` is an alias for `H2BC5` (source: https://www.ncbi.nlm.nih.gov/gene/3017)
* `HIST1H3D` is an alias for `H3C4` (source: https://www.ncbi.nlm.nih.gov/gene/8351)

Each of these HGNC names is present in our table:

```{r}
map_table |>
  dplyr::filter(gene_symbol %in% c("ADA2", "H2BC5", "H3C4"))
```


## Conclusions

We'll need to recode the following symbols in our objects before passing them into `EPIC`.

* `CXorf36` should be changed to `DIPK2B`
* `CECR1` should be changed to `ADA2`
* `HIST1H2BC` should be changed to `H2BC5`
* `HIST1H3D` should be changed to `H3C4`


## Session info

```{r}
sessionInfo()
```
