---
title: "Explore EPIC results"
author: Stephanie J. Spielman
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: show
---

The goal of this notebook is to explore the `EPIC` results obtained with both of its references `TRef` or `BRef`. 
These two distinct references estimate the proportions of the following cell types (note they use these strings, not ontologies), where I have bolded the cell types they share:

- The `TRef` reference (7 + other):
  - **`Bcells`**    
  - `CAFs` (cancer-associated fibroblasts)
  - **`CD4_Tcells`**
  - **`CD8_Tcells`**
  - `Endothelial`
  - `Macrophages`
  - **`NKcells`**
  - `otherCells`
- The `BRef` reference (6 + other):
  - **`Bcells`**   
  - **`CD4_Tcells`**
  - **`CD8_Tcells`**
  - `Monocytes`
  - `Neutrophils`
  - **`NKcells`**
  - `otherCells`
          
`EPIC` was run on bulk samples from five different projects with the high-level diagnoses shown in the table below.
Based on the biology of these groups of tumors, we plan to analyze each with the reference indicated in the table.


| Project     | High-level diagnosis | `EPIC` Reference |
|-------------|----------------------|------------------|
| SCPCP000001 | High-grade glioma    | `TRef`           |   
| SCPCP000002 | Low-grade glioma     | `TRef`           |   
| SCPCP000006 | Wilms tumor          | `BRef`           |   
| SCPCP000009 | CNS tumors           | `TRef`           |   
| SCPCP000017 | Osteosarcoma         | `BRef`           |   


## Setup

```{r setup, message = FALSE}
renv::load()

library(ggplot2)
theme_set(theme_bw())
```


### Paths

```{r paths}
results_dir <- here::here("analysis", "bulk-deconvolution", "results")

epic_files <- list.files(
  path = results_dir,
  full.names = TRUE,
  pattern = "*-epic.tsv$"
)
project_ids <- stringr::str_split_i(basename(epic_files), pattern = "-", i = 1)
epic_files <- setNames(epic_files, project_ids)
```

### Prepare input data

First, we'll read in the `EPIC` results and subset to only the reference of interest for each. 
We'll retain the `reference` column in the final data frame since it will help with visualization.

```{r}
tref_projects <- c("SCPCP000001", "SCPCP000002", "SCPCP000009")
bref_projects <- c("SCPCP000006", "SCPCP000017")

# Read in epic result TSVs and combine into single data frame
epic_raw_df <- epic_files |>
  purrr::map(\(file) readr::read_tsv(file, show_col_types = FALSE)) |>
  purrr::list_rbind(names_to = "project_id") 

# Filter based on projects and references
epic_tref <- epic_raw_df |>
  dplyr::filter(
    project_id %in% tref_projects,
    reference == "TRef"
  )
 
epic_bref <- epic_raw_df |>
  dplyr::filter(
    project_id %in% bref_projects,
    reference == "BRef"
  )

# Create final epic_df with only references of interest for each project
epic_df <- dplyr::bind_rows(
  epic_tref, 
  epic_bref
) |>
  # add column to support interpreting figure panels with projects, diagnosis, and EPIC reference
 dplyr::mutate(
   project = dplyr::case_match(
      project_id, 
      "SCPCP000001" ~ "01: High-grade glioma (TRef)", 
      "SCPCP000002" ~ "02: Low-grade glioma (TRef)", 
      "SCPCP000006" ~ "06: Wilms tumor (BRef)", 
      "SCPCP000009" ~ "09: CNS tumors (TRef)", 
      "SCPCP000017" ~ "17: Osteosarcoma (BRef)", 
    ), 
   # arrange in order of reference, not project
   project = forcats::fct_relevel(project, "09: CNS tumors (TRef)", after = 2)
  ) |>
  dplyr::select(project, sample_id, epic_celltype, reference, fraction)

head(epic_df)
```



## Visualization

### Proportions of cell types

Considering all categories of cells, what proportions of cell types are represented across samples?


```{r fig.height=8, fig.width=12}
# Next, we'll define a vector of colors to use for plots which ensures that shared cell types have the same colors
# hex codes are (mostly) from Okabe-Ito
color_map <- c(
  # shared cell types
  "Bcells" = "#E69F00", 
  "CD4_Tcells" = "#56B4E9", 
  "CD8_Tcells" = "#009E73", 
  "NKcells" =  "#F0E442",
  # reference-specific cell types
  "CAFs" = "#0072B2",
  "Endothelial" = "#D55E00",
  "Macrophages" = "#CC79A7",
  "Monocytes" = "grey30", 
  "Neutrophils" = "grey50",
  "otherCells" = "grey95"
)  
celltype_levels <- c(names(color_map))
  
ggplot(epic_df) + 
 aes(
   x = sample_id, 
   y = fraction, 
   fill = forcats::fct_relevel(epic_celltype, celltype_levels)
  ) +
  geom_col() + 
  facet_wrap(vars(project), scales = "free_x", nrow = 2) + 
  scale_y_continuous(expand = c(0,0)) +
  scale_fill_manual(values = color_map, name = "cell type") +
  theme(
    legend.position = "bottom", 
    axis.text.x = element_blank()
  )
```

In most samples, the vast majority of cells are `otherCells`, although there are two samples which are estimated to have ~100% cells classified; it is very likely that cells in these samples (and possibly more samples) are incorrectly classified.

We also see these trends:

* For the `TRef` projects:
  * Macrophages are most commonly observed in `SCPCP000001` samples
  * Endotheliel cells tend to dominate `SCPCP000001` and `SCPCP000002`
* For the `BRef` projects:
  * CD4 T-cells and monocytes are more prevalent in `SCPCP000017` compared to `SCPCP000006`
  * CD8 T-cells dominate `SCPCP000006` cell types

Let's have a quick look at the proportion of classified cells on its own:

```{r}
epic_classified_df <- epic_df |>
  dplyr::filter(epic_celltype == "otherCells") |>
  dplyr::mutate(fraction_classified = 1 - fraction) 

ggplot(epic_classified_df) +
  aes(x = project, y = fraction_classified) + 
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = rel(0.9)))

```
In the vast majority of samples, fewer than 25% of cells are classified, with the fewest cells classified in Wilms Tumor. 


Notably, across samples, there appear to be very few B-cells inferred, and when they are inferred the frations are extremely small (to the point that we might essentially consider them to be 0).
Let's look at this a bit more closely, zooming in on the section of the plot with most of the data:

```{r, fig.width = 10}
epic_bcell_df <- epic_df |>
  dplyr::filter(epic_celltype == "Bcells") 

ggplot(epic_bcell_df) +
  aes(x = project, y = fraction) + 
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = rel(0.9))) + 
  ggforce::facet_zoom(ylim = c(0, 0.02)) 
```


### Number of identified cell types

How many cell types are represented in each sample?
For this, we'll make a histogram of the number of cell types identified per sample (regardless of their proportion).
Since, as it seems, EPIC generally doesn't estimate 0's, we'll instead provide a low threshold value of `1e-4` (artbitrarily chosen as low but not _too_ low!).

Note that the maximum number of cell types is 7 for `TRef` and 6 for `BRef` (not including other).
```{r}
celltype_counts_df <- epic_df |>
  dplyr::filter(epic_celltype != "otherCells") |>
  dplyr::group_by(project, sample_id) |>
  dplyr::summarize(num_celltypes = sum(fraction > 1e-4))
```


```{r fig.height=4, fig.width=10}
ggplot(celltype_counts_df) + 
  aes(x = num_celltypes) + 
  geom_histogram(center = 0, binwidth = 1) + 
  scale_x_continuous(breaks = scales::breaks_pretty()) +
  facet_wrap(vars(project), nrow = 1) + 
  ggtitle("Number of cell types inferred in individual samples")
```  
We can also make an associated summary table:

```{r}
celltype_counts_df |>
  dplyr::group_by(project) |>
  dplyr::summarize(
    min_celltypes = min(num_celltypes), 
    median_celltypes = median(num_celltypes), 
    max_celltypes = max(num_celltypes), 
  )
```

Overall, we see that most `TRef`-analyzed cell types have 6 cell types, and most `BRef`-analyzed samples have 4. 

Let's look at how many samples contain each cell type:
Again, this only considers cell type presence and absence at a `1e-4` threshold.

```{r}
# calculate number of celltypes per sample
samples_per_celltype_df <- epic_df |>
  dplyr::filter(epic_celltype != "otherCells") |>
  dplyr::mutate(celltype_present = fraction > 1e-4) |>
  dplyr::group_by(epic_celltype, project) |>
  dplyr::summarize(n_samples = sum(celltype_present)) 

# convert number to proportion
samples_per_celltype_df <- epic_df |>
  dplyr::filter(epic_celltype != "otherCells") |>
  dplyr::count(project) |>
  dplyr::right_join(samples_per_celltype_df) |>
  dplyr::mutate(prop_samples = n_samples / n)
```

```{r fig.height=6, fig.width=8, message = FALSE}
ggplot(samples_per_celltype_df) + 
  aes(
    x = forcats::fct_reorder(epic_celltype, prop_samples),
    y = prop_samples) + 
  geom_col() +
  facet_wrap(vars(project), nrow = 2, scales = "free_x") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = rel(0.8))) + 
  labs(
    x = "cell type",
    title = "Proportion of samples with each cell type"
  )
```
The main new trend we can pull out from here is that `NKcells` are rarely observed in samples from `SCPCP000001` and `SCPCP000006`, and are virtually unobserved in `SCPCP000002` and `SCPCP000009`.



## Conclusions

`EPIC` appears to be a more conservative method and, frequently, cell types present in the reference are not present in the samples even though we might have _a priori_ expectations they are present. 
There are also samples which were virtually entirely classified, but these inferences are not unlikely to be spurious.
The next step will be to compare these inferences with `quanTIseq` inferences.

## Session info

```{r}
sessionInfo()
```